Community-Conditioned Language Models

### 2020-11-12 paper/paper.tex
revisions

\newcommand\BibTeX{B\textsc{ib}\TeX}

@@

To condition the model, we add a \emph{community embedding} parameter to the models.
This parameter is concatenated with the hidden layer of the sequence encoder,
for some layer $c \leq n$, and passed through a linear layer 
which projects the resulting vector back to the origial hidden layer size.
For $c = n$, the output of this linear layer is passed directly to the softmax function,
just as the final hidden layer of the sequence encoder is in other models.
For $c=0$, the community embedding is concatenated with the token embedding.
For this reason, we set the hidden size of the sequence encoder 
and the size of the token to be embedding equal for all models.

@@

We also note that the conditioned models perform differently across different communities---\
for some communities, the test perplexity of the conditioned models is a lot lower (Table~\ref{tab:model-ppls},
whereas for others it is the same or even slightly higher.
\bn{TODO: talk about this in terms of mutual information instead?}
As an anecdotal observation, we note 
that communities where conditioning has the least effect
tend to be more organized around more ``general interest'' topics.\jp{Is there a way to quantify general interest topics? Community size in terms of number of users?}
This makes sense intuitively, since communities with more niche topics
would tend to have more specialized vocabulary and speech patterns.

The pattern of how much the perplexity falls across different communities
is similar across architectures, but with some differences.
It could be interesting to further investigate these differences in future work,
since it could reveal differences in the kind of linguistic variation the different
model architectures capture.

@@

- Two different use of the model
1. ``absolute'' perplexity
2. latent ability to recognize a community

% However, we are not so much interested in the ability of the CCLM to precisely model the language used in a community.

We now turn our interest to the Bayesian classifier induced by a CCLM, which we call a language-model-based community classifier
(LMCC). Indeed, the purpose of an LMCC is to estimate the probability $P(m|c_i)$ that a message $m$
belong in a community $c_i$. Using the Bayes theorem, we can turn a
CCLM into a LMCC: the probability that a given message $m$
belongs to a community $c_i$ can be calculated as follows:
\[P(c=c_i | m) = P(m | c_i)\frac {P(m)} {P(c=c_i)}\]
In the above,
$P(c=c_i)$ is the frequency of community $c_i$ in the dataset, and
$P(m | c_i)$ is given by the CCLM. $P(m)$ is the absolute
probability of a message, which can be computed as
\[P(m) = \frac 1 {\sum_i P(c=c_i) P(m|c=c_i ) }\].

We can characterise the quality the LMCC by a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i | m)$ for a random message $m$ in
community $c_j$.  That is:
\[C_{ij} = average_{m \in Messages(c_j)}(P(c=c_i | m))\].
\jp{Is this test set or training set?}

@@

TODO: why don't we use F scores?

While the confusion matrices provide a lot of insight, they are
somewhat \emph{too} fine-grained. For this reason, we will be
interested in confusing the messages from community $c_j$ are to the
LMCC. For this purpose, we consider the perplexity of the LMCC for a
message coming from community $j$, defined as the exponential of the
entropy of the distribution \(D_j\), which is the (discrete)
probability distribution of predicted community for a random message from
$c_j$. $D_j$ is simply the $j$th row in \(C\), and so we obtain the formula:
\[Ppl_j = e^{-\sum_i C_{ij} log(C_{ij})}\]

To recap, \(Ppl_j\) is a measure of the capability (or rather
inability) of the model to correctly identify a message as coming from
a certain community $c_j$.

@@

\begin{figure}
  \pplscatter{lstm-3-1_cclm}{lstm-3-1_lmcc}
  \caption{%
    Mean CCLM (x-axis) and LMCC (y-axis) perplexity per community (Pearson's r = \num{0.30}, p = \num{0.0444}).
    Shown here for the LSTM model with $c = 1$, but other models exhibit similar behavior.
  }
  \label{fig:cclm_lmcc_ppl}
\end{figure}

We can then analyse the correlation between \(Ppl^M_j\) for various models $M$.

\begin{figure}
  \pplscatter{lstm-3-1_lmcc}{transformer-3-3_lmcc}
  \caption{%
    Mean CCLM perplexity per community for the best LSTM ($c = 1$) and transformer ($c = 3$) models
    (Pearson's r = \num{0.99} p < \num{0.0001}).
  }
  \label{fig:lmcc_ppl}
\end{figure}

>>>>

\newcommand\Ppl{\mathsf{Ppl}}
\newcommand\Ind{I}
\newcommand\BibTeX{B\textsc{ib}\TeX}

@@

To condition the model, we add a \emph{community embedding} parameter to the models.
This parameter is concatenated with the hidden layer of the sequence encoder,
for some layer $c \leq n$, and passed through a linear layer 
which projects the resulting vector back to the origial hidden layer size.
For $c = n$, the output of this linear layer is passed directly to the softmax function,
just as the final hidden layer of the sequence encoder is in other models.
For $c=0$\jp{it's unfortunate that the letter $c$ is clashing. To Fix!}, the community embedding is concatenated with the token embedding.
For this reason, we set the hidden size of the sequence encoder 
and the size of the token to be embedding equal for all models.

@@

We also note that the conditioned models perform differently across different communities---\
for some communities, the test \jp{and training?}perplexity of the conditioned models is a lot lower (Table~\ref{tab:model-ppls},
whereas for others it is the same or even slightly higher.
\bn{TODO: talk about this in terms of mutual information instead?}
As an anecdotal observation, we note 
that communities where conditioning has the least effect
tend to be more organized around more ``general interest'' topics.\jp{Is there a way to quantify general interest topics? Community size in terms of number of users?}
This makes sense intuitively, since communities with more niche topics
would tend to have more specialized vocabulary and speech\jp{can you say speech for written language?} patterns.

The pattern of how much the perplexity falls across different communities
is similar across architectures, but with some differences.\jp{Either say something specific here or move the whole paragraph to future work.}
It could be interesting to further investigate these differences in future work,
since it could reveal differences in the kind of linguistic variation the different
model architectures capture.

@@

The perplexity of a CCLM tells is a measure of linguistic complexity,
but it is not a measure of how linguistically \emph{specific} the
community is. However, the CCLM induces a latent 
a latent classifier induced by a CCLM, which we call a language-model-based community classifier
(LMCC). Indeed, using the Bayes theorem, we can turn a
CCLM into a LMCC: the probability that a given message $m$
belongs to a community $c_i$ can be calculated as follows:
\[P(c=c_i | m) = P(m | c_i)\frac {P(m)} {P(c=c_i)}\]
In the above,
$P(c=c_i)$ is the frequency of community $c_i$ in the dataset, and
$P(m | c_i)$ (the probability that a message $m$
belongs in a community $c_i$. ) is estimated by the CCLM. $P(m)$ is the absolute
probability of a message, which can be computed as
\[P(m) = \frac 1 {\sum_i P(c=c_i) P(m|c=c_i ) }\].

We can characterise the quality the LMCC by a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i | m)$ for a random message $m$ in
community $c_j$.  That is:
\[C_{ij} = average_{m \in Messages(c_j)}(P(c=c_i | m))\].
As a sanity check, we observe that the diagonal dominates the matrix: on average the LMCC picks the right community for a message.

\jp{Is this test set or training set?}

@@

However, we want to measure the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use the perplexity metric
(again), which is has stronger information-theoretic grounds.
However, this time, the perplexity corresponds to the difficulty of
identifying a community, not predicting a word.
Technically, we consider the average perplexity of the LMCC for a
message coming from community $j$, defined as the exponential of the
entropy of the distribution \(D_j\), which is the (discrete)
probability distribution of predicted community for a random message from
$c_j$. $D_j$ is simply the $j$th row in \(C\). In sum, we have the formula:
\[\Ind_j = e^{-\sum_i C_{ij} log(C_{ij})}\]

We call \(\Ind_j\) the (linguistic) indiscernibility of the community
$c_j$, and it is a measure of how unsure a model is about which
community a message belongs to --- when the message in question is
actually coming from a community $c_j$.

@@

Indiscernibility is a robust measure: it is almost perfectly
correlated for various CCLM (transformer, lstm and various $c$) $M$.
\jp{other r's}
\begin{figure}
  \pplscatter{lstm-3-1_lmcc}{transformer-3-3_lmcc}
  \caption{%
    Mean CCLM perplexity per community for the best LSTM ($c = 1$) and transformer ($c = 3$) models
    (Pearson's r = \num{0.99} p < \num{0.0001}).
  }
  \label{fig:lmcc_ppl}
\end{figure}

However we observe that the CCLM perplexity is only weakly correlated with the
indiscernibility to the corresponding latent LMCC
(\cref{fig:cclm_lmcc_ppl}).

\begin{figure}
  \pplscatter{lstm-3-1_cclm}{lstm-3-1_lmcc}
  \caption{%
    Mean linguistic perplexity (x-axis) and community indiscernibility
    (y-axis) per community (Pearson's r = \num{0.30}, p =
    \num{0.0444}).  Shown here for the LSTM model with $c = 1$, but
    other models exhibit similar behavior.  }
  \label{fig:cclm_lmcc_ppl}
\end{figure}
